Assessing Phonon Properties of Shape Memory Alloys using Graph Neural Network Potentials

### 2024-01-07 src/tex/ms.tex
Mods to rules, env, phonon calc, writing ms.

% Define document class
\documentclass[preprint]{elsarticle}
\usepackage{showyourwork}

@@

 % Main body with filler text
\section{Introduction}
\label{sec:intro}
Shape memory alloys derive their unique displacive phase transformations from the dynamic instablity that occurs during the application of temperature, strain, or stress. The origin of this instability occurs due to softening or symmetry breaking of the phonon modes in the phase. Therefore, improved understanding of the phonon characteristics of the different phases that can occur for a given SMA is particular important. One of the challenges is in performing such calculations, which typically employ density functional theory in order to determine the atomic forces, which in turn are used to construct the dynamical matrix. However, the use of classical interatomic potentials can alleviate some of this computational burden. \par

@@

The subsequent step once the featureization process is completed is to determine a suitable neural network archetcture which can learn the target property or properties given the feature inputs. In the case of atomistic simulations the target is prediction of the potential energy surface and corresponding gradients (i.e., forces). Two of the most early attempts to do this were the works of \citet{Bartok2010} and \citet{Thompson2015}, whereby projection onto higher-dimensional basis, such as bispectrum componenets of local neighbor density, is learned in order to describe the atomic environments. The energy per atom is then determined by these learned features by employing linear regression or non-linear Gaussian process regression. \par

While these early attempts were successfully, they could only include structural details to learn new features from. Futhermore, the training process was typically restricted to a limited number of elemental constieunts and therefore for each chemical system and phases of interest one needs to provide a suitable training dataset. To move beyond such limitations, groups began using graph neural network (GNN) architectures and message passing scheme to represent the atomic structures as nodes and edges \cite{Reiser2022}. This enabled two distinctions, first being that the nodes and edges corresponding to the atoms and bonds could include features, and secondly the learning could be done over a large structural and compositional space enabling a generalization\footnote{Typically these are referred to as Universal interatomic potentials or forcefields since they in theory generalize well to any chemical system or phase.}. The use of GNNs is efficient in representing structure-composition-property inputs to additional NN layers for predicting potential energy surfaces and gradients and was initially pursued in the CGCNN, MEGNet, and ALIGNN potentials \cite{Choudhary2021,Chen2019,Xie2018}. The differences between the approaches are predominately in atomic and bond features, nearest-neighbor graph, and specific NN archetcture. \par

The training of these early GNN utlized large density functional theory calculation repositories and mostly tuned the learning cost function to ground-state structure and energy. This resulted in mean-absolute-errors and other measures of performance that could be improved upon \cite{Riebesell2023}. More recently there have been significant improvements in both training data used and the GNN architecture that show promise for truly approaching a general interatomic potential suitable for static and dynamic atomistic calculations of multicomponents and multiphases. If succesfull, identifying and designing new potential SMA materials using GNN interatomic potentials could have tremendous impact on technology. Therefore, the primary goal is to provide a qualitative assement of predictions of the equations of state, phonon bandstructure, and elastic constants for various phases of NiTi and PtTi using three GNN potentials intended for general use: M3GNet, CHGNet, and MACE \cite{Chen2022,Deng2023,Batatia2022}. We note that the \citet{Riebesell2023} provides a summarized performance over these GNN using the materials project database, but does not specifically address SMA and phonon characteristics. \par

@@

\subsection{Workflow}
\label{sec:workflow}
To promote reproducible research and align with findable, accessible, interoper-able and reusable (FAIR) principles\cite{Walsh2024}, we utilize \showyourwork python workflow\cite{Luger2021} and several open source python packages. Instructions for reproducing calculation data, figures, tables, and manuscript are provided in appendix~\ref{sec:appx_reproduce} and corresponding github repo\footnote{\url{https://github.com/stefanbringuier/SMA_Phonons_GNNIP}}. \par

@@

\subsubsection{Phonon Bandstructures}
Phonon calculations are performend using finite-difference method via small displacement (i.e., harmonic approximation) of atoms in a reference cell \cite{Alfe2009}. In the small displacement method, atoms in the primitive cell are slightly displaced to calculate forces on other atoms, constructing the force constant matrix, which, when used with the dynamical matrix, allows computing phonon frequencies at any q-vector in the Brillouin zone. This requires typically requires the use of a supercell of the original unit cell. In order to ensure convergence of this method, one needs to sweep different supercell sizes and atomic displacements. This process is done for each interatomic potential and structure. The supercells used in this work  have a repeat in any one direction ranging from $5-8$ and depends on the crystal system. The atomic displacement ranges from $0.03-0.26$\AA\,and was found to strongly depend on interatomic potential and crystal system. Details on the settings are provided in the appendix~\ref{sec:phonon_calc_config}. \par

@@

The EAM and MEAM potentials are

@@

The M3GNet has been implemented in the MatGL framework and we utlize the \texttt{M3GNet-MP-2021.2.8-PES} pretrained model.

@@

The thermodynamic equation of state (EOS) is a standard assement of the phase stability of materials at a give condition. In the context of atomic calculations the equation of state can be determined at ground-state (i.e., 0K) to asses the phase stability of the different crystals. The equation of state can be represented as a pressure vs. volume or more commonly as the energy per atom  vs the ratio of the isotropically strained volume to that of the equilibrium volume. One criteria for a interatomic potential is that it accurately refelcts the EOS curves for different phases. This entails the construction of a convex hull which represents the minimum energy curve accross the structures for the volume strain conditions. \par

The EOS curves for the \ce{NiTi} structures \textit{B2}, \textit{B19}, \textit{B19'}, \textit{B33}, and \textit{Pbcm}  for the different potentials studied in this work are shown in fig.~\ref{fig:eos}. The two EAM potentials, show that at 0 strain ($V/V_o = 1.0$) both the \textit{B19P} and \textit{B33} phases are stable. I believe this is the first reporting of the \textit{B33} phase for these two fittings of the EAM potential for \ce{NiTi}. Its important to note that these potentials have been developed to capture the displacive phase transformation between \textit{B19P} to \textit{B2}. This is shown in fig.~\ref{fig:eos}(a,b) with the stability indicated upon compressive isotropic strain. A mininum appears around $\approx 0.9$. At tensile isotropic strains the \textit{B19P} phase is stable up till around $1.18$ where the \textit{Pbcm} phase shows stability. An interesting note is that these two fittings of the EAM potential show a double well curve for the \textit{B2} phase. \par

@@

\section*{Reproducibilty and Data Availability}
This study was carried out using the reproducibility software \href{https://github.com/showyourwork/showyourwork}{\showyourwork}\cite{Luger2021}, which leverages continuous integration to programmatically download the data from \href{https://zenodo.org/}{zenodo.org}(if needed), create the figures, and compile the manuscript. Each figure caption contains two links: one to the dataset stored on zenodo used in the corresponding figure, and the other to the script used to make the figure (at the commit corresponding to the current build of the manuscript). The git repository associated to this study is publicly available at \url{https://github.com/stefanbringuier/SMA_Phonons_GNNIP} and allows anyone to re-build the entire manuscript.\par

@@

\begin{figure}[!htp]                                                                                      
    \script{AppendixBZ.py}                                                                           
    \begin{centering}                                                                                    
      \includegraphics[width=0.5\linewidth]{figures/B2_BrillouinZonePointsSampled.png} 
      \vspace{1mm}                                                                                       
      \includegraphics[width=0.5\linewidth]{figures/B19_BrillouinZonePointsSampled.png}
      \vspace{1mm}
      \includegraphics[width=0.5\linewidth]{figures/B19P_BrillouinZonePointsSampled.png}
      \vspace{1mm}
      \includegraphics[width=0.5\linewidth]{figures/BCO_BrillouinZonePointsSampled.png}
      \vspace{1mm}                                                                                       
      \caption{The irreducible Brillouin zones for \textit{B2}(\#221), \textit{B19}(\#51), \textit{B19'}(\#11), and \textit{BCO}(\#63) crystal structures}                                                               \label{fig:ibz}                                                                          
    \end{centering}                                                                                      
\end{figure}

>>>>

% Define document class
\documentclass[preprint,colorlinks=true,linkcolor=black,citecolor=black]{elsarticle}
\usepackage{showyourwork}

@@

 % Main body with filler text
\section{Introduction}
\label{sec:intro}

The design and development of shape memory alloys (SMA) via \textit{in-silico} methods would provide great value in advancing structural, medical, and other technologies. In order to enable such activity would require sampling from very large structural and compositional spaces and evaluating material properties indicative of displacive phase transformation phenomena. The use of density functional theory calculations to obtain properties such as equations of state, phonon disperisons, and elastic constants, can be done, however, such calculations are computationally demanding and are potentially prohibitive for smaller groups trying to develop new SMAs. To alleviate such restriction, one can resort to classical interatomic potentials which can greatly improve computational cost in  evaluating the energies and forces more. However, the immediate challenge is to whether a suitable interatomic potential exist given a chemical system, and moreover, how transferable is it to different compositions and phases. \par

@@

The subsequent step once the featureization process is completed is to determine a suitable neural network (NN) architecture which can learn the target property or properties given the feature inputs. In the case of atomistic simulations the target is prediction of the potential energy surface and corresponding gradients (i.e., forces). Two of the most early attempts to do this were the works of \citet{Bartok2010} and \citet{Thompson2015}, whereby projection onto higher-dimensional basis, such as bispectrum components of local neighbor density, is learned in order to describe the atomic environments. The energy per atom is then determined by these learned features by employing linear regression or non-linear Gaussian process regression. We note that these are just two of the early examples but similar potentials have been developed in earlier, in parallel, or recently by various groups \cite{Behler2007,Wang2018,Deringer2019,Lee2019}. \par

While these early attempts using ML models were successfully, they may only include limited representations of structural and/or chemical details to learn new features from which could limit transferability. Futhermore, the training process was typically restricted to a limited number of elemental constieunts of the target material system. As a consequence,  for each chemical system and phases of interest one needs to provide a suitable training dataset. To move beyond such limitations, groups began using graph neural network (GNN) architectures and message passing scheme to represent the atomic structures as nodes and edges \cite{Reiser2022}. This enabled two distinctions, first being that the nodes and edges corresponding to the atoms and bonds could include features, and secondly the learning could be done over a large structural and compositional space enabling a generalization\footnote{Typically these are referred to as Universal interatomic potentials or forcefields since they in theory generalize well to any chemical system or phase.}. The use of GNNs is efficient in representing structure-composition-property inputs to additional NN layers for predicting potential energy surfaces and gradients and was initially pursued in the CGCNN, MEGNet, and ALIGNN potentials \cite{Choudhary2021,Chen2019,Xie2018}. The differences between the approaches are predominately in atomic and bond features, nearest-neighbor graph, and specific NN archetcture. \par

The training of these early GNN utlized large density functional theory calculation repositories and mostly tuned the learning cost function to ground-state structure and energy. This resulted in mean-absolute-errors and other measures of performance that could be improved upon \cite{Riebesell2023}. More recently there have been significant improvements in both training data used and the GNN architecture that show promise for truly approaching a general interatomic potential suitable for static and dynamic atomistic calculations of multicomponents and multiphases. If succesfull, identifying and designing new potential SMA materials using GNN interatomic potentials could have tremendous impact on technology. Therefore, the primary goal is to provide a qualitative assessment of predictions of the equations of state, phonon bandstructure, and elastic constants for various phases of NiTi and PtTi using three GNN potentials intended for general use: M3GNet, CHGNet, and MACE \cite{Chen2022,Deng2023,Batatia2022}. We note that the \citet{Riebesell2023} provides a summarized performance over these GNN using the materials project database, but does not specifically address SMA and phonon characteristics. \par

@@

\subsection{Workflow}
\label{sec:workflow}
To promote reproducible research and align with findable, accessible, interoper-able and reusable (FAIR) principles\cite{Walsh2024}, we utilize \showyourwork python workflow\cite{Luger2021} and several open source python packages. Instructions for reproducing calculated data, figures, tables, and manuscript are provided in appendix~\ref{sec:appx_reproduce} and corresponding github repo\footnote{\url{https://github.com/stefanbringuier/SMA_Phonons_GNNIP}}.\par

@@

\subsubsection{Phonon Bandstructures}

Shape memory alloys derive their unique displacive phase transformations from the dynamic instability that occurs during the application of temperature, strain, or stress. The origin of this instability occurs due to softening or symmetry breaking of the phonon modes in the phase. Therefore, improved understanding of the phonon characteristics of the different phases that can occur for a given SMA is particular important. One of the challenges is in performing such calculations, which typically employ density functional theory (DFT) in order to determine the atomic forces, which in turn are used to construct the dynamical matrix. However, classical interatomic potentials can alleviate the computational burden. \par

Phonon calculations are performed using finite-difference method via small displacement (i.e., harmonic approximation) of atoms in a reference cell \cite{Alfe2009}. In the small displacement method, atoms in the primitive cell are slightly displaced to calculate forces on other atoms, constructing the force constant matrix, which, when used with the dynamical matrix, allows computing phonon frequencies at any q-vector in the Brillouin zone. This requires typically requires the use of a supercell of the original unit cell. In order to ensure convergence of this method, one needs to sweep different supercell sizes and atomic displacements. This process is done for each interatomic potential and structure. The supercells used in this work  have a repeat in any one direction ranging from $5-8$ and depends on the crystal system. The atomic displacement ranges from $0.03-0.26$\AA\,and was found to strongly depend on interatomic potential and crystal system. Details on the settings are provided in the appendix~\ref{sec:phonon_calc_config}. \par

@@

The EAM and MEAM potentials were developed to go beyond pair interatomic potentials to treat point and extended defects in metal and alloys systems. The motivation was taken from the physical results of  density functional theory and tight-binding methods, namely to add to the pair potentials a energy functional which depends on the local electron density which in turn is affected by the atomic environment. These type of interatomic potentials are broadly referred to as EAM or MEAM, with the later included direction (i.e. angle) dependent interactions. \par

The functional form using during fitting proceed depends on the complexity of the material system but the total energy is typically written as,

\begin{equation}
  \label{eq:eam}
  E = \sum_{i\neq j}^N V_{ij}(r_{ij}) + \sum_i^N F_i\left(\rho(r_i)\right)
\end{equation}

where $V_{ij}$ is the pair potential, $F_i$, the functional, and $\rho$ the density function that depends on the environment around $r_i$. The MEAM potential extends this by including three-body terms in the $\rho$ to capture features such as covalent bonding. The EAM and MEAM potentials have been particular successfully in predicting structure and properties of metals and alloys, including defect formation/energies, phase transitions, grain boundary stability/dynamics. One particular advantage of both EAM and MEAM, is their analytical or tabulated forms allow for quick numerical evaluation, with the primary reason being that the energies and forces on a atom are accurately determined within a relatively small interaction volume (i.e., 3-4 \AA interaction cutoff).  \par

For application of EAM to shape memory alloys, and in particular NiTi, \citet{Mutter2010} initially devised an Finnis-Sinclair style EAM potential to capture the structural phase transitions between \textit{B19'} and \textit{B2} phases. It was latter shown by \cite{Zhong2011} that some adjustments to the EAM parameters was needed to properly capture the finite-temperature displacement dynamics for the martensitic transition \cite{Gur2017}. More recently \citet{Ko2015} developed a second nearest neighbor(2NN) MEAM parameterization of NiTi that demonstrated very good predictive performance and improvements over ealier potentials.

@@

The use of ML potentials for SMA has been demonstrated by \citet{Tang2022}. In their work, they utilized the DeePMD-Kit\cite{Wang2018} neural network framework to parameterize a descriptor matrix  from atomic configurations that preserves permutations and  symmetries. The descriptors are then which feed into a multilayer perceptron to calculate energies and forces. The neural network interatomic potential demonstrated good accuracy of ground-state properties with respect to DFT results and showed agreement with experiments regarding temperature, stress, and defect induced martensitic transformation. \paragraphmark

In the work of \citet{Tang2022} ground-state predictions for \textit{B2}, \textit{B19'}, \textit{B19}, and \textit{BCO}\footnote{Referred to as \textit{B33} prototype in \citet{Tang2022}} were provided and showed good agreement. However, we note that deviation from Ni-Ti 1:1 stoichiometry or addition of a third species was not investigated nor the inclusion of additional alloying species. \par

The pre-trained GNN interatomic potentials explored in this work have been parameterized over a large fraction of the interatomic potential ($>$ 80 elements) and therefore are not restricted to specific chemistries or stoichiometries. They training data also includes various phases for a given chemical system. This permits great flexibility. We do not that fine-tuning, the process of adjusting GNN weights and biases using additional training data, can be performed with the GNN studied in this work. However, this is out of the scope for this manuscript, given resources and time needed to generate such training data for SMAs. \par

The M3GNet is a GNN that utilizes a 3-body bond-update procedure to more accurately capture bond-order environment changes. The M3GNet has been implemented in the MatGL framework and we utilize the \texttt{M3GNet-MP-2021.2.8-PES} pretrained model.

@@

The EOS curves for the \ce{NiTi} structures \textit{B2}, \textit{B19}, \textit{B19'}, \textit{BCO} for the different potentials studied in this work are shown in fig.~\ref{fig:eos}. The two EAM potentials, show that at 0 strain ($V/V_o = 1.0$) both the \textit{B19P} and \textit{B33} phases are stable. I believe this is the first reporting of the \textit{B33} phase for these two fittings of the EAM potential for \ce{NiTi}. Its important to note that these potentials have been developed to capture the displacive phase transformation between \textit{B19P} to \textit{B2}. This is shown in fig.~\ref{fig:eos}(a,b) with the stability indicated upon compressive isotropic strain. A mininum appears around $\approx 0.9$. At tensile isotropic strains the \textit{B19P} phase is stable up till around $1.18$ where the \textit{Pbcm} phase shows stability. An interesting note is that these two fittings of the EAM potential show a double well curve for the \textit{B2} phase. \par

@@

\section*{Reproducibility and Data Availability}
This study was carried out using the reproducibility software \href{https://github.com/showyourwork/showyourwork}{\showyourwork}\cite{Luger2021}, which leverages continuous integration to programmatically download the data from \href{https://zenodo.org/}{zenodo.org}(if needed), create the figures, and compile the manuscript. Each figure caption contains two links: one to the dataset stored on zenodo used in the corresponding figure, and the other to the script used to make the figure (at the commit corresponding to the current build of the manuscript). The git repository associated to this study is publicly available at \url{https://github.com/stefanbringuier/SMA_Phonons_GNNIP} and allows anyone to re-build the entire manuscript.\par

@@

\begin{figure}[!htp]
  \centering
  \begin{subfigure}[b]{0.45\linewidth}
      \includegraphics[width=\linewidth]{figures/B2_BrillouinZonePointsSampled.png}
      \caption{}
      \label{fig:B2}
  \end{subfigure}
  \hfill
  \begin{subfigure}[b]{0.45\linewidth}
      \includegraphics[width=\linewidth]{figures/B19_BrillouinZonePointsSampled.png}
      \caption{}
      \label{fig:B19}
  \end{subfigure}
  \vspace{1mm}
  \begin{subfigure}[b]{0.45\linewidth}
      \includegraphics[width=\linewidth]{figures/B19P_BrillouinZonePointsSampled.png}
      \caption{}
      \label{fig:B19P}
  \end{subfigure}
  \hfill
  \begin{subfigure}[b]{0.45\linewidth}
      \includegraphics[width=\linewidth]{figures/BCO_BrillouinZonePointsSampled.png}
      \caption{}
      \label{fig:BCO}
  \end{subfigure}
  \caption{The irreducible Brillouin zones for \textit{B2}(\#221), \textit{B19}(\#51), \textit{B19'}(\#11), and \textit{BCO}(\#63) crystal structures}
  \label{fig:ibz}
\end{figure}

\pagebreak
